\bf Eye Fitting Straight Lines in the Modern Era

### 2022-08-03 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
knitted manuscript revisions

We all use statistical graphics, but how do we know that the graphics we
use are communicating properly? When creating a graphic, we must
consider the design choices most effective for conveying the intended
result. For instance, we may decide to highlight the relationship
between two variables in a scatter-plot by including a trend line, or
adding color to highlight clustering \citep{vanderplas2017clusters}.
These design choices require that we understand the perceptual and
visual biases that come into play when creating graphics, and as
graphics are evaluated visually, we must use human testing to ground our
understanding in empiricism.

Much of the research on the perception of visual features in charts has
been conducted in psychophysics and tests for accuracy and quantitative
comparisons when understanding a plot. \citet{cleveland1984graphical}
conducted a series of cognitive tasks designed to establish a hierarchy
of visual components for making comparisons. For example, it is more
effective to display information on an \(x\) or \(y\) axis rather than
using color in order to reduce the visual effort necessary to make
numerical comparisons. \citet{cleveland1985graphical} found that
assessing the position of points along an axis is easier than
determining the slope of a line. Other studies focus on the viewers
ability to perceive the strength of the relationship between \(x\) and
\(y\) coordinates in a scatter-plot. For instance, when the data appear
dense, viewers tend to overestimate the magnitude of the correlation
coefficient \citep{cleveland1982variables, lauer1989density}.
\citet{cleveland1993visualizing} provides an argument for displaying
cyclical patterns with an aspect ratio which sets the curve close to
45\(^{\circ}\). \citet{kosslyn2006graph} examines how Gestalt principles
of perceptual organization are instrumental in extracting data from a
chart. For example, data points that are closer in space are more likely
to be perceived as grouped \citep{ciccione2020grouping}, and vertical
and horizontal lines are easier to discriminate than oblique ones
\citep{appelle1972perception}. The results of these cognitive tasks
provide some consistent guidance for chart design; however, other
methods of visual testing can further evaluate design choices and help
us understand cognitive biases related to the evaluation of statistical
charts.

@@

We need human testing of graphics in order to draw broad conclusions,
develop guidelines for graphical design, and improve graphical
communication. Studies might ask participants to identify differences in
graphs, read information off of a chart accurately, use data to make
correct real-world decisions, or predict the next few observations. All
of these types of tests require different levels of use and manipulation
of the information being presented in the chart. Early research studies
considered graphs from a psychological perspective
\citep{spence1990visual, lewandowsky1989perception}, testing
participants' abilities to detect a stimulus or a difference between two
stimuli. Psychophysical methods have been used to test graphical
perception, as in \citet{vanderplas2015signs}, which used the method of
adjustment to estimate the magnitude of the impact of the sine illusion.
However, there are more modern testing methods that have been developed
since the heyday of psychophysics.

One major development in statistical graphics which led to more advanced
testing methods is Wilkinson's Grammar of Graphics
\citep{wilkinson2013grammar}. The grammar of graphics serves as the
fundamental framework for data visualization with the notion that
graphics are built from the ground up by specifying exactly how to
create a particular graph from a given data set. Visual representations
are constructed through the use of ``tidy data'' which is characterized
as a data set in which each variable is in its own column, each
observation is in its own row, and each value is in its own cell
\citep{wickham2016r}. Graphics are viewed as a mapping from variables in
a data set (or statistics computed from the data) to visual attributes
such as the axes, colors, shapes, or facets on the canvas in which the
chart is displayed. Software, such as Hadley Wickham's ggplot2
\citep{wickham2011ggplot2}, aims to implement the framework of creating
charts and graphics as the grammar of graphics recommends.

Combining the grammar of graphics with another tool for statistical
graphics testing, the statistical lineup, yields a method for evaluating
graphical design choices. \citet{buja2009statistical} introduced the
lineup protocol to provide a framework for inferential testing. Through
experimentation, methods such as the lineup protocol allow researchers
to conduct studies geared at understanding human ability to conduct
tasks related to the perception of statistical charts such as
differentiation, prediction, estimation, and extrapolation
\citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
The advancement of graphing software provides the tools necessary to
develop new methods of testing statistical graphics. While these testing
methods are excellent, there is one particular subset of statistical
graphics testing methods which we intend to develop further in this
paper: assessing graphics by fitting statistical models ``by eye''.

@@

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points
\citep{finney1951subjective, mosteller1981eye}. Common methods of
fitting trends by eye involved maneuvering a string, black thread, or
ruler until the fit is suitable, then drawing the line through the set
of points. Recently, \citet{ciccione2021can} conducted a comprehensive
set of studies investigating human ability to detect trends in graphical
representations using physical adjustment and manipulation methods.

@@

\citet{mosteller1981eye} sought to understand the properties of least
squares and other computed lines by establishing one systematic method
of fitting lines by eye. Participants were asked to fit lines by eye to
four scatter-plots using an 8.5 x 11 inch transparency with a straight
line etched completely across the middle. A latin square design with
packets of the set of points stapled together in four different
sequences was used to determine if there is an effect of order of
presentation. It was found that order of presentation had no effect and
that participants tended to fit the slope of the principal axis (error
minimized orthogonally, both horizontal and vertical, to the regression
line) over the slope of the least squares regression line (error
minimized vertically to the regression line). These results support
previous research on ``ensemble perception'' indicating the visual
system can compute averages of various features in parallel across the
items in a set
\citep{chong2003representation, chong2005statistical, van2011rapid}.

In \citet{ciccione2021can}, participants were asked to judge trends,
estimate slopes, and conduct extrapolation. To estimate slopes,
participants were asked to report the slope of the best-fitting
regression line using a track-pad to adjust the tilt of a line on the
screen. Results indicated the slopes participants reported were always
in excess of the ideal slopes, both in the positive and in the negative
direction, and those biases increase with noise and with number of
points. This supports the results found in \citet{mosteller1981eye} and
suggest that participants might use Deming regression
\citep{deming1943statistical, linnet1998performance, martin2000general},
which minimizes the Euclidean distance of points from the line, when
fitting a line to a noisy scatter-plot.

While not explicitly intended for perceptual testing, in 2015, the New
York Times introduced an interactive feature, called You Draw It
\citep{aisch_cox_quealy_2015, buchanan_park_pearce_2017, katz_2017}.
Readers were asked to input their own assumptions about various metrics
and compare how these assumptions relate to reality. The New York Times
team utilizes Data Driven Documents (D3) \citep{bostock2011d3} that
allows readers to predict these metrics through the use of drawing a
line on their computer screen with their computer mouse. After the
reader has completed drawing the line, the actual observed values are
revealed and the reader may check their estimated knowledge against the
actual reported data. While this interactive feature is designed to get
readers to confront their own intuitions about data in the news, we feel
that the interactivity of this method may be useful for the purpose of
graphical testing and measuring the patterns humans see in data.

In this paper, we establish `You Draw It', adapted from the New York
Times feature, as a tool for graphical testing. Our visual system is
naturally built to look for structure and identify patterns. For
instance, points going down from left to right indicates a negative
correlation between the \(x\) and \(y\) variables. Our research is
intended to implement the `You Draw It' feature as a way to measure the
patterns we see in data. We validate the `You Draw It' method by
replicating the less technological study conducted by
\citet{mosteller1981eye}. Based on previous research surrounding
``ensemble perception,'' we hypothesize that visual regression tends to
mimic principle component regression rather than an ordinary least
squares regression. In order to assess this hypothesis, we introduce a
method for statistically modeling the participant drawn lines using
generalized additive mixed models.

@@

Participants were recruited through through Twitter, Reddit, and direct
email in May 2021. A total of 35 individuals completed 119 unique `You
Draw It' task plots; data collected from all completed `You Draw It'
task plots were included in the analysis. All participants had normal or
corrected to normal vision and signed an informed consent form. The
experimental tasks took approximately 15 minutes to complete. While this
study does utilize a convenience sample, as this is primarily a
perceptual task, previous results have found few differences between
expert and non-expert participants in this context
\citep{vanderplas2015spatial}. These data were collected to validate
this method of graphical testing, with the hopes of providing a new tool
to assess graphical perception interactively. Participants completed the
experiment on their own computers in an environment of their choosing.
The experiment was conducted and distributed through a Shiny application
\citep{shinyPkg} found
\href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{here}.

@@

In the study, participants were shown an interactive scatter-plot
(\cref{fig:ydi-stimuli}) along with the prompt, ``Use your mouse to fill
in the trend in the yellow box region.'' The yellow box region moved
along as the user drew their trend-line, providing a visual cue which
indicates where the user still needed to complete a trend line. After
the entire domain had been visually estimated or predicted, the yellow
shaded region disappeared, indicating the participant had completed the
task. Data Driven Documents (D3), a JavaScript-based graphing framework
that facilitates user interaction, was used to create the `You Draw It'
visual. In order to allow for user interaction and data collection, we
integrated the D3 visual into Shiny using the \texttt{r2d3} package.
While the interface is highly customized to this particular task, we
hope to generalize the code and provide a Shiny widget in an R package
soon.

@@

All data processing was conducted in R statistical software. A total of
\(N = 30\) points \((x_i, y_i), i = 1,...N\) were generated for
\(x_i \in [x_{min}, x_{max}]\) where \(x\) and \(y\) have a linear
relationship. Data were simulated based on a linear model with additive
errors: \begin{align}
y_i & = \beta_0 + \beta_1 x_i + e_i \\
\text{with } e_i & \sim N(0, \sigma^2). \nonumber
\end{align}

Model equation parameters, \(\beta_0\) and \(\beta_1\), were selected to
reflect the four data sets (F, N, S, and V) used in
\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). We obtained
\(\beta_0\) from the point-slope equation of a line using the mean of
the generated \(x\) values and the predefined \(y\) value at \(\bar x\),
denoted \(y_{\bar x}\). Parameter choices F, N, and S simulated data
across a domain of 0 to 20. Parameter choice F produced a trend with a
positive slope and a large variance while N had a negative slope and a
large variance. In comparison, S showed a trend with a positive slope
and a small variance while V yielded a steep positive slope with a small
variance over the domain of 4 to 16. \cref{fig:eyefitting-simplot}
illustrates an example of simulated data for all four parameter choices
intended to reflect the trends in \citet{mosteller1981eye}. Aesthetic
design choices were made consistent across each of the interactive `You
Draw It' task plots. The y-axis range extended 10\% beyond (above and
below) the range of the simulated data points to allow for users to draw
outside the simulated data set range and avoid anchoring their lines to
the corners of the plot.

@@

\caption{\label{tab:eyefitting-parameters}Designated model equation parameters for simulated data.}
\centering
\begin{tabular}[t]{cccc}
\toprule
Parameter Choice & $y_{\bar{x}}$ & $\beta_1$ & $\sigma$\\
\midrule
S & 3.88 & 0.66 & 1.30\\
F & 3.90 & 0.66 & 1.98\\
V & 3.89 & 1.98 & 1.50\\
N & 4.11 & -0.70 & 2.50\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices.}\label{fig:eyefitting-simplot}
\end{figure}

@@

This experiment was conducted as part of a larger study of the
perception of log and linear scales; for simplicity, we focused on the
study design and methods related to the current study. Each scatter-plot
was the graphical representation of a data set that was generated
randomly and independently for each participant at the start of the
experiment. Participants in the study were shown two `You Draw It'
practice plots in order to train participants in the skills associated
with executing the task - in particular, the responsiveness of the
applet requires that participants draw a line at a certain speed,
ensuring that all of the evenly spaced points along the hand-drawn line
are filled in. During the practice session, participants were provided
with instruction prompts accompanied by a .gif and a practice plot.
Instructions guided participants to start at the edge of the yellow box,
to make sure the yellow shaded region was moving along with their mouse
as they drew, and that they could draw over their already drawn line.
Practice plots were then followed by four `You Draw It' task plots
associated with the current study. The order of the task plots was
randomly assigned for each individual in a completely randomized design.

@@

We compared the participant drawn line to two regression lines
determined by ordinary least squares (OLS) regression and regression
based on the principal axis (PCA). \cref{fig:ols-vs-pca-example}
illustrates the difference between an OLS regression line which
minimizes the vertical distance of points from the line and a regression
line based on the principal axis which minimizes the Euclidean distance
of points (orthogonal) from the line.

Due to the randomness in the data generation process, the actual slope
of the linear regression line fit through the simulated points could
differ from the predetermined slope. Therefore, we fit an OLS regression
to each scatter-plot to obtain estimated parameters
\(\hat\beta_{0,OLS}\) and \(\hat\beta_{1,OLS}\). Fitted values,
\(\hat y_{k,OLS}\), were then obtained every 0.25 increments across the
domain from the OLS regression equation,
\(\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k\)., for
\(k = 1, ..., 4 x_{max} +1\). The regression equation based on the
principal axis was determined by using the \texttt{princomp} function in
the stats package in base R to obtain the rotation of the coordinate
axes from the first principal component (direction which captures the
most variance). The estimated slope, \(\hat\beta_{1,PCA}\), was
determined by the ratio of the axis rotation in \(y\) and axis rotation
in \(x\) of the first principal component with the y-intercept,
\(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
using the mean of the simulated points, \((\bar x_i, \bar y_i)\). Fitted
values, \(\hat y_{k,PCA}\), were then obtained every 0.25 increment
across the domain from the PCA regression equation,
\(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).

@@

For each participant, the final data set used for analysis contained
\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PCA}\)
for parameter choice \(i = 1,2,3,4\), j = \(1,...N_{participant}\), and
\(x_{ijk}\) value for increment \(k = 1, ...,4 x_{max} + 1\). Using both
a linear mixed model and a generalized additive mixed model, comparisons
of vertical residuals in relation to the OLS fitted values
(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were made
across the domain. \cref{fig:eyefitting-example-plot} displays an
example of all three fitted trend lines for parameter choice F.

@@

\caption{Illustrates the data associated with and collected for one `You Draw It' task plot. Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
\end{figure}

@@

Using the \texttt{lmer} function in the lme4 package \citep{lme4}, a
linear mixed model (LMM) was fit separately to the OLS residuals and PCA
residuals, constraining the fit to a linear trend. Parameter choice,
\(x\), and the interaction between \(x\) and the parameter choice were
treated as fixed effects with a random participant effect included to
account for variation due to participant. The LMM equation for each fit
(OLS and PCA) is given by: \begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation} \noindent where

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn \(y\) value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\) value
\item
  \(\hat y_{ijk,fit}\) is the fitted \(y\) value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\) value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted \(y\)
  values for the \(i^{th}\) parameter choice, \(j^{th}\) participant,
  and \(k^{th}\) increment of \(x\) value corresponding to either the
  OLS or PCA fit
\item
  \(\gamma_0\) is the overall intercept
\item
  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
  V, N) on the intercept
\item
  \(\gamma_1\) is the overall slope for \(x\)
\item
  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
\item
  \(x_{ijk}\) is the \(x\) value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
  the \(j^{th}\) participant's characteristics
\item
  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
\end{itemize}

>>>>

We all use statistical graphics, but how do we know that the graphics we
use are communicating properly? When creating a graphic, we must
consider the design choices most effective for conveying the intended
result. For instance, we may decide to highlight the relationship
between two variables in a scatterplot by including a trend line, or
adding color to highlight clustering \citep{vanderplas2017clusters}.
These design choices require that we understand the perceptual and
visual biases that come into play when creating graphics, and as
graphics are evaluated visually, we must use human testing to ground our
understanding in empiricism.

Much of the research on the perception of visual features in charts has
been conducted in psychophysics and tests for accuracy and quantitative
comparisons when understanding a plot. \citet{cleveland1984graphical}
conducted a series of cognitive tasks designed to establish a hierarchy
of visual components for making comparisons. For example, it is more
effective to display information on an \(x\) or \(y\) axis rather than
using color in order to reduce the visual effort necessary to make
numerical comparisons. \citet{cleveland1985graphical} found that
assessing the position of points along an axis is easier than
determining the slope of a line. Other studies focused on the viewers'
ability to perceive the strength of the relationship between \(x\) and
\(y\) coordinates in a scatterplot. For instance, when the data appear
dense, viewers tend to overestimate the magnitude of the correlation
coefficient \citep{cleveland1982variables, lauer1989density}.
\citet{cleveland1993visualizing} provided an argument for displaying
cyclical patterns with an aspect ratio which sets the curve close to
45\(^{\circ}\). \citet{kosslyn2006graph} examined how Gestalt principles
of perceptual organization are instrumental in extracting data from a
chart. For example, \citet{ciccione2020grouping} conducted a study to
support data points located closer together are more likely to be
perceived as the same group and \citet{appelle1972perception} found that
it is easier to discriminate vertical and horizontal lines than oblique
lines. The results of these cognitive tasks provided some consistent
guidance for chart design; however, other methods of visual testing can
further evaluate design choices and help us understand cognitive biases
related to the evaluation of statistical charts.

@@

We need human testing of graphics in order to draw broad conclusions,
develop guidelines for graphical design, and improve graphical
communication. Studies might ask participants to identify differences in
graphs, read information off of a chart accurately, use data to make
correct real-world decisions, or predict the next few observations. All
of these types of tests require different levels of use and manipulation
of the information being presented in the chart. Early research studies
considered graphs from a psychological perspective
\citep{spence1990visual, lewandowsky1989perception}, testing
participants' abilities to detect a stimulus or a difference between two
stimuli. Psychophysical methods have been used to test graphical
perception, as in \citet{vanderplas2015signs}, which used the method of
adjustment - a technique which requires participants to alter a changing
stimulus to match a given constant stimuli \citep{gescheider_1997}- to
estimate the magnitude of the impact of the sine illusion. However,
there are more modern testing methods that have been developed since the
heyday of psychophysics.

One major development in statistical graphics which led to more advanced
testing methods is Wilkinson's Grammar of Graphics
\citep{wilkinson2013grammar}. The Grammar of Graphics serves as the
fundamental framework for data visualization with the notion that
graphics are built from the ground up by specifying exactly how to
create a particular graph from a given data set. Visual representations
are constructed through the use of ``tidy data'' which is characterized
as a data set in which each variable is in its own column, each
observation is in its own row, and each value is in its own cell
\citep{wickham2016r}. Graphics are viewed as a mapping from variables in
a data set (or statistics computed from the data) to visual attributes
such as the axes, colors, shapes, or facets on the canvas in which the
chart is displayed. Software, such as Hadley Wickham's ggplot2
\citep{wickham2011ggplot2}, aims to implement the framework of creating
charts and graphics as the Grammar of Graphics recommends.

Combining the Grammar of Graphics with another tool for statistical
graphics testing, the statistical lineup, yields a method for evaluating
graphical design choices. \citet{buja2009statistical} introduced the
lineup protocol to provide a framework for inferential testing. A
statistical lineup is a plot consisting of smaller panels where the
viewer is asked to identify the target panel containing the real data
from among a set of decoy null plots which display data under the
assumption there is no relationship. If the viewer can identify the
target panel randomly embedded within the set of null panels, this
suggests that the real data is visually distinct from data generated
under the null model. Through experimentation, methods such as the
lineup protocol allow researchers to conduct studies geared at
understanding human ability to conduct tasks related to the perception
of statistical charts such as differentiation, prediction, estimation,
and extrapolation
\citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
The advancement of graphing software provides the tools necessary to
develop new methods of testing statistical graphics. While these testing
methods are excellent, there is one particular subset of statistical
graphics testing methods which we intend to develop further in this
paper: assessing graphics by fitting statistical models ``by eye''.

@@

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points
\citep{unwin1988eyeballing, finney1951subjective, mosteller1981eye}.
Common methods of fitting trends by eye involved maneuvering a string,
black thread, or ruler until the fit is suitable, then drawing the line
through the set of points. Recently, \citet{ciccione2021can} conducted a
comprehensive set of studies investigating human ability to detect
trends in graphical representations using physical adjustment and
manipulation methods.

@@

\citet{mosteller1981eye} sought to understand the properties of least
squares and other computed lines by establishing one systematic method
of fitting lines by eye. Participants were asked to fit lines by eye to
four scatterplots using an 8.5 x 11 inch transparency with a straight
line etched completely across the middle. A latin square design with
packets of the set of points stapled together in four different
sequences was used to determine if there is an effect of order of
presentation. It was found that order of presentation had no effect and
that participants tended to fit the slope of the principal axis (PA)
(error minimized orthogonally, both horizontal and vertical, to the
regression line) over the slope of the least squares regression line
(error minimized vertically to the regression line).

In \citet{ciccione2021can}, participants were asked to judge trends,
estimate slopes, and conduct extrapolation. To estimate slopes,
participants were asked to report the slope of the best-fitting
regression line using a track-pad to adjust the tilt of a line on the
screen. Results indicated the slopes participants reported were always
in excess of the ideal slopes, both in the positive and in the negative
direction, and those biases increase with noise and with number of
points. This supports the results found in \citet{mosteller1981eye} and
suggest that participants might use Deming regression
\citep{deming1943statistical, linnet1998performance, martin2000general},
which minimizes the Euclidean distance of points from the line, when
fitting a line to a noisy scatterplot.

While not explicitly intended for perceptual testing, in 2015, the New
York Times introduced an interactive feature, called `You Draw It'
\citep{aisch_cox_quealy_2015, buchanan_park_pearce_2017, katz_2017}.
Readers were asked to input their own assumptions about various metrics
and compare how these assumptions relate to reality. The New York Times
team utilizes Data Driven Documents (D3) \citep{bostock2011d3} that
allows readers to predict these metrics through the use of drawing a
line on their computer screen with their computer mouse. After the
reader has completed drawing the line, the actual observed values are
revealed and the reader may check their estimated knowledge against the
actual reported data. While this interactive feature is designed to get
readers to confront their own intuitions about data in the news, we feel
that the interactivity of this method may be useful for the purpose of
graphical testing and measuring the patterns humans see in data.

In this paper, we establish `You Draw It', adapted from the New York
Times feature, as a new tool for graphical testing. Our visual system is
naturally built to look for structure and identify patterns. For
instance, points going down from left to right indicates a negative
correlation between the \(x\) and \(y\) variables. Our research is
intended to implement the `You Draw It' feature as a way to measure the
patterns we see in data. The graphical testing method used in this study
differs from prior methods found in \citet{mosteller1981eye} and
\citet{ciccione2021can} by allowing participants to freely draw
estimated trend lines - a method which extends nicely to a nonlinear
setting. We validate the `You Draw It' method by replicating the less
technological study conducted by \citet{mosteller1981eye}. In
\protect\hyperlink{methods}{Section 2} we describe our participant
sample, the graphical task to be completed, and the data generation
process and study design. \protect\hyperlink{results}{Section 3}
describes the participant data collected and shares results from the
analyses of the data using mixed models. Overall conclusions and
discussion of results are presented in
\protect\hyperlink{conclusion-discussion}{Section 4} with extensions to
the current work suggested in \protect\hyperlink{future-work}{Section
5}. The RShiny applet to complete the study, participant data used for
analysis, and code to replicate the analysis can be found in the
\protect\hyperlink{supplementary-material}{Supplementary Material}.
Based on previous research, we hypothesize that visual regression tends
to mimic regression based on the principal axis rather than an ordinary
least squares regression. In order to assess this hypothesis, we
introduce a method for statistically modeling the participant drawn
lines using generalized additive mixed models (GAMM). While the focus of
this paper is to share the results from the validation study which uses
the new `You Draw It' method to evaluate visually fitted linear trends
to statistical regression results, the intent of this work is to set the
foundation and demonstrate the strength of the combination of the `You
Draw It' method and GAMM analyses for testing statistical graphics and
to extend the use of the method beyond the linear setting.

@@

Participants were recruited through through Twitter, Reddit, and direct
email in May 2021. A total of 35 individuals completed 131 unique `You
Draw It' task plots. Data were collected as a part of a pilot study
meant to test the applet; therefore, either voluntary participant
dropout or disconnection from a server not designed to accommodate large
magnitudes of participants resulted in missing plots in our data set for
analysis. All participants had normal or corrected to normal vision and
signed an informed consent form. The experimental tasks took
approximately 15 minutes to complete. As this is a pilot study,
participants from Twitter and Reddit pages related to data visualization
voluntarily completed the study and likely have an interest in fields
related to statistics and want to help advance research in graphics.
While this study does utilize a convenience sample, as this is primarily
a perceptual task, previous results have found few differences between
expert and non-expert participants in this context
\citep{vanderplas2015spatial}. These data were collected to validate
this method of graphical testing, with the hopes of providing a new tool
to assess graphical perception interactively. Participants completed the
experiment on their own computers in an environment of their choosing.
The experiment was conducted and distributed through a Shiny application
\citep{shinyPkg} found at
\href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{emily-robinson.shinyapps.io/you-draw-it-validation-applet}.

@@

In the study, participants were shown an interactive scatterplot
(\cref{fig:ydi-stimuli}) along with the prompt, ``Use your mouse to fill
in the trend in the yellow box region.'' The yellow box region moved
along as the user drew their trend-line, providing a visual cue which
indicates where the user still needed to complete a trend line. After
the entire domain had been visually estimated or predicted, the yellow
shaded region disappeared, indicating the participant had completed the
task. Data Driven Documents (D3), a JavaScript-based graphing framework
that facilitates user interaction, was used to create the `You Draw It'
visual. In order to allow for user interaction and data collection, we
integrated the D3 visual into Shiny using the \texttt{r2d3} package
\citep{r2d3}. While the interface is highly customized to this
particular task, we hope to generalize the code and provide a Shiny
widget in an R package soon.

@@

All data processing was conducted in R software environment for
statistical computing and graphics \citep{Rsoftware}. A total of
\(N = 30\) points \((x_i, y_i), i = 1,...,N\) were generated for
\(x_i \in [x_{min}, x_{max}]\) where \(x\) and \(y\) have a linear
relationship. Data were simulated based on the point-slope form of a
linear model with additive errors: \begin{align}
y_i = \beta_1(x_i-\bar{x}) + y_{\bar{x}} + e_i \\
\text{with } e_i & \sim N(0, \sigma^2). \nonumber
\end{align}

Model equation parameters, \(\beta_1\), \(y_{\bar{x}}\), and parameter
choice letter names (F, S, V, N), were selected to reflect the four data
sets used and labeled in \citet{mosteller1981eye}
(\cref{tab:eyefitting-parameters}). The mean of the generated \(x\)
values and the predefined \(y\) value at \(\bar x\), denoted
\(y_{\bar x}\) were used in the point-slope equation of a line.
Parameter choices S, F, and N simulated data across a domain of 0 to 20.
Parameter choice F produced a trend with a positive slope and a large
variance while N had a negative slope and a large variance. In
comparison, S showed a trend with a positive slope and a small variance
while V yielded a steep positive slope with a small variance over the
domain of 4 to 16. \cref{fig:eyefitting-simplot} illustrates an example
of simulated data for all four parameter choices intended to reflect the
trends in \citet{mosteller1981eye}. Aesthetic design choices were made
consistent across each of the interactive `You Draw It' task plots. The
y-axis range extended 10\% beyond (above and below) the range of the
simulated data points to allow for users to draw outside the simulated
data set range and avoid anchoring their lines to the corners of the
plot.

@@

\caption{\label{tab:eyefitting-parameters}Designated model equation parameters for simulated data.}
\centering
\begin{tabular}[t]{ccccc}
\toprule
Parameter Choice & $y_{\bar{x}}$ & $\beta_1$ & $\sigma$ & Domain\\
\midrule
F & 3.90 & 0.66 & 1.98 & (0,20)\\
S & 3.88 & 0.66 & 1.30 & (0,20)\\
V & 3.89 & 1.98 & 1.50 & (4,16)\\
N & 4.11 & -0.70 & 2.50 & (0,20)\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{Example of simulated data points displayed in a scatterplot illustrating the trends associated with the four selected parameter choices.}\label{fig:eyefitting-simplot}
\end{figure}

@@

This experiment was conducted as part of a larger study of the
perception of log and linear scales; for simplicity, we focused on the
study design and methods related to the current study. Each data set was
generated randomly and independently for each participant at the start
of the experiment and mapped to a scatterplot. Participants in the study
were shown two `You Draw It' practice plots in order to train
participants in the skills associated with executing the task - in
particular, the responsiveness of the applet requires that participants
draw a line at a certain speed, ensuring that all of the evenly spaced
points along the hand-drawn line are filled in. During the practice
session, participants were provided with instruction prompts accompanied
by a .gif and a practice plot. Instructions guided participants to start
at the edge of the yellow box, to make sure the yellow shaded region was
moving along with their mouse as they drew, and that they could draw
over their already drawn line. Practice plots were then followed by one
of each of the four `You Draw It' task plots associated with the current
study (S, F, V, and N). The order of the task plots was randomly
assigned for each individual in a completely randomized design.

@@

We compared the participant drawn line to two regression lines
determined by ordinary least squares (OLS) regression and regression
based on the principal axis (PA). \cref{fig:ols-vs-pca-example}
illustrates the difference between an OLS regression line which
minimizes the vertical distance of points from the line and a regression
line based on the PA which minimizes the Euclidean distance of points
(orthogonal) from the line.

Due to the randomness in the data generation process, the actual slope
of the linear regression line fit through the simulated points could
differ from the predetermined slope. Therefore, we fit an OLS regression
to each scatterplot to obtain estimated parameters \(\hat\beta_{0,OLS}\)
and \(\hat\beta_{1,OLS}\). Fitted values, \(\hat y_{k,OLS}\), were then
obtained every 0.25 increment across the domain from the OLS regression
equation,
\(\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k\), for
\(k = 1, ..., 4 x_{max} +1\). The PA regression slope,
\(\hat\beta_{1,PA}\), and y-intercept, \(\hat\beta_{0,PA}\), were
determined using the \texttt{mcreg} function in the \texttt{mcr} package
in R \citep{mcr_pkg} which implements Deming regression (equivalent to a
regression based on the slope of the first principal axis). Fitted
values, \(\hat y_{k,PA}\) were then obtained every 0.25 increment across
the domain from the PA regression equation,
\(\hat y_{k,PA} = \hat\beta_{0,PA} + \hat\beta_{1,PA} x_k\), for
\(k = 1, ..., 4 x_{max} +1\).

@@

For each participant, the final data set used for analysis contained
\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PA}\)
for parameter choice \(i = 1,2,3,4\), j = \(1,...,N_{participant}\), and
\(x_{ijk}\) value for increment \(k = 1, ...,4 x_{max} + 1\). Using both
a linear mixed model and a generalized additive mixed model, comparisons
of vertical residuals in relation to the OLS fitted values
(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PA fitted
values (\(e_{ijk,PA} = y_{ijk,drawn} - \hat y_{ijk,PA}\)) were made
across the domain. \cref{fig:eyefitting-example-plot} displays an
example of all three fitted trend lines for parameter choice F.

@@

\caption{Illustrates the data associated with and collected for one `You Draw It' task plot. Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
\end{figure}

@@

The `You Draw It' method does not restrict participants to draw a
straight line as other methods would, such as using a ruler. Instead,
participants are allowed to freely draw a line with potential curvature.
Using the \texttt{lmer} function in the lme4 package \citep{lme4}, a
linear mixed model (LMM) was fit separately to the OLS residuals and PA
residuals, emulating the effect of constraining participants to draw a
linear trend. Both fixed and random parameter estimates in the LMM were
determined by optimizing the restricted maximum likelihood (REML)
through penalized least squares. Parameter choice, \(x\), and the
interaction between \(x\) and the parameter choice were treated as fixed
effects with a random participant effect included to account for
variation due to participant. The LMM equation for each fit (OLS and PA)
is given by: \begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation} \noindent where

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn \(y\) value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\) value
\item
  \(\hat y_{ijk,fit}\) is the fitted \(y\) value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\) value corresponding to either the OLS or PA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted \(y\)
  values for the \(i^{th}\) parameter choice, \(j^{th}\) participant,
  and \(k^{th}\) increment of \(x\) value corresponding to either the
  OLS or PA fit
\item
  \(\gamma_0\) is the overall intercept
\item
  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
  V, N) on the intercept
\item
  \(\gamma_1\) is the overall slope for \(x\)
\item
  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
\item
  \(x_{ijk}\) is the \(x\) value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
  the \(j^{th}\) participant's characteristics
\item
  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
\end{itemize}
